\documentclass{article}

%settings
\usepackage[utf8]{inputenc}
\usepackage[explicit]{titlesec}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{braket}
\usepackage[margin=1.0in]{geometry}
\usepackage{bbold}
\usepackage{fancyhdr}
\usepackage{fancyvrb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\pagestyle{fancy}
\fancyhead[L]{\leftmark}
\fancyhead[R]{\thepage}

\title{POL486: Networks in International Politics}
\author{Shamel Bhimani}
\date{Fall 2025}

\begin{document}

\maketitle

\tableofcontents
\newpage

    \section{Introduction to Network Analysis}
    \subsection{Olga V. Chyzh. Network analysis in international relations. In Cameron G. Thies, editor, Handbook of International Relations, pages 158–170. Edward Elgar Publishing, 2025.}
    \subsubsection{Networks in International Relations}
    \noindent \textbf{Core Premise:} International politics is inherently
networked. Actors (states, organizations, individuals) are interconnected
nodes, and their relationships (alliances, trade, conflict) are ties.\\

    \noindent \textbf{Significance of Networks:}
    \begin{itemize}
        \item Membership in international clubs (e.g., NATO, EU, WTO) offers
        security, prestige, and economic benefits.
        \item Exclusion can lead to insecurity and foreign policy revisionism.
        \item Network embeddedness affects policy options and resource access.
    \end{itemize}

    \noindent \textbf{Historical Context:} IR scholars have long recognized
the networked nature of global politics, but network analysis provided the
specific tools to better align theory with empirical evidence.\\

    \noindent \textbf{Critique of Traditional IR Research:}
    \begin{itemize}
        \item Network scholars criticized previous IR research, particularly
        the \textbf{dyadic design}, for its failure to account for
        interdependence among actors.
        \item The assumption of independence in dyadic analysis can lead to
        confounding bias, attributing effects to incorrect causes. For
        example, the US-Japan and US-South Korea relationships influence the
        Japan-South Korea relationship.
    \end{itemize}

    \subsubsection{Methodological and Theoretical Contributions}
    \noindent \textbf{Primary Contribution:} Network analysis offered a way
to measure previously unmeasurable concepts like system polarity, social
power, and prestige. It also generated new research questions about
connectivity.\\

    \noindent \textbf{Three Main Research Approaches:}
    \begin{itemize}
        \item \textbf{Global Network Properties:} Studies focus on
        properties of the entire network, such as density of
        fractionalization, to explain outcomes like conflict and cooperation.
        \begin{itemize}
            \item [$1$.]\textbf{Example:} Maoz (2006) used network analysis to
            create a theory-informed measure of international system
            polarization, a concept previously hard to operationalize.
            \item \textbf{Example:} Cruz, Labonne, and Querubin (2020) found
            that greater fractionalization (power divided among more clans)
            in local kinship networks in the Philippines was associated with
            better public goods provision.
        \end{itemize}
        \item [$2$.] \textbf{Actor Positions:} This approach analyzes the
        positions of individual actors within a network to understand power
        and influence.
        \begin{itemize}
            \item \textbf{Brokerage/Gate-keeping Power:} Held by actors
            connecting otherwise discontinued clusters.
            \item \textbf{Network Centrality:} Used as a proxy for concepts
            like country prestige (Renshon 2016) or an organization's
            agenda-setting power (Carpenter 2011).
        \end{itemize}
        \item[$3$.] \textbf{Overlapping Membership:} This apporach examines
        how overlapping memberships reinforce each other's effects.
        \begin{itemize}
            \item \textbf{Example:} Parkisnon (2013) showed that sustaining
            an insurgency depends on the overlap between military networks
            and personal networks (kinship, friendship).
            \item \textbf{Example:} Eldredge and Shannon (2022) found that
            countries with high membership overlap in inter-governmental
            organizations are more likely to object to each other's human
            rights treaty reservations.
        \end{itemize}
    \end{itemize}

    \subsubsection{The Debate and Normalization of Network Analysis in IR}
    \noindent \textbf{The `Us-vs-Them' Debate:} Early proponents of network
analysis adopted a provocative framing, creating divisions within IR.\\

    \noindent \textbf{Points of Resistance:} Critics argued that the
traditional dyadic approach had not impeded major theoretical advances (
e.g., the democratic peace) and that research designs should be tailored to
the specific question, rather than assuming interdependence as the default.\\

    \noindent \textbf{Mainstreaming the Approach (c. 2016):}
    \begin{itemize}
        \item An exchange in \textit{International Relations Quarterly}
        between proponents and critics marked a key moment.
        \item A special issue on networks in
        \textit{Journal of Peace Research} showcased the breadth of
        applications.
        \item The field of political methodology quickly welcomed and
        published inferential network analysis research.
        \item Subsequently, network research began appearing in top
        disciplinary journals, at major conferences, and in university
        curricula.
    \end{itemize}

    \subsubsection{Current Research and Future Directions}
    \noindent \textbf{Addressing Endogeneity:} Developing tools to separate
actor-level effects from network-level effects (e.g., democracy vs. clique
size in trade).\\

    \noindent \textbf{Flexible Conceptualization:} Re-evaluating the unit of
analysis,
such as treating alliances themselves as nodes to study action-reaction
processes.\\

    \noindent \textbf{Expanding Scope:} Applying network analysis to
subnational and transnational levels, including rebel groups, NGOs, and
political elites.\\

    \noindent \textbf{Social Media Data:} Utilizing vast, inherently
networked data from social media to study mobilization, censorship, and
misinformation.\\

    \noindent \textbf{Future Directions:}
    \begin{itemize}
        \item \textbf{Develop IR-Specific Theories:} Move beyond borrowing
        sociological theories to build network theories tailored to IR's
        unique actors and assumptions (e.g., anthropomorphizing states).
        \item \textbf{Model Hierarchical Networks:} Incorporate asymmetrical
        and hierarchical relationships, not just horizontal ones between
        equal actors.
        \item \textbf{Integrate Casual Inference:} Bridge network analysis
        with experimental and quasi-experimental methods to test network
        predictions more rigorously.
        \item \textbf{Model Co-evolution:} Better theorize and moel the
        endogenous relationship where actor characteristics are both a cause
        and an effect of their network ties.
    \end{itemize}


    \section{Network Centrality}
    \subsection{John F. Padgett and Christopher K. Ansell. Robust action and the rise of the Medici, 1400-1434. American Journal of Sociology, 98(6):1259–1319, 1993.}
    \subsubsection{Introduction and Core Argument}

    \noindent \textbf{Central Claim:} The rise of Medicean political
control (1400-1434) in Florence, leading to the Renaissance state, was
driven by \textbf{network disjunctures} within the elite that the Medici
alone spanned.\\

    \noindent \textbf{Methodology:} To understand state formation, one must
move beyond formal institutions, groups, and goals to the
\textbf{relational substrata} of people's actual lives.
\textbf{Ambiguity and heterogeneity}, not planning and self-interest, are
the raw materials of powerful states and persons.\\

    \noindent \textbf{Key Concept:} \textit{Robust Action}, Cosimo de
Medici's (1380-1464) control style, characterized by
\textbf{multivocal identity as `sphinx'}, which harnessed power from these
`network holes' and resolved the inherent contradiction between `judge' and
`boss' in organizations.

    \subsubsection{Core Concepts Explained}

    \noindent \textbf{Political Centralization and the Judge/Boss Contradiction}
    \begin{itemize}
        \item State-building involves centralizing power, a contradictory
        process requiring both reproduction (rules creating roles,
        interests, collective action patterns) and control (others'
        interactions serving one's interests).
        \item The contradiction: A founder cannot be both an impartial
        `judge' (legitimacy through non-self-interest) and a controlling
        `boss' (direct intervention undermines legitimacy).
    \end{itemize}

    \noindent \textbf{Robust Action:}
    \begin{itemize}
        \item \textbf{Definition:} A style of control where single actions
        can be coherently interpreted from multiple perspectives
        simultaneously (multivocality), serve as moves in many `games' at
        once, and blur public/private motivations.
        \item \textbf{Mechanism:} Leads to `Rorschach blot identities',
        where others attribute their own distrinctive identity to the ego (
        Cosimo).
        \item \textbf{Goal:} Maintaining \textbf{flexible opportunism} and
        discretionary options in unpredictable futures, rather than pursuing
        specific, fixed goals. This involves \textbf{positional play:}
        maneuvering opponents into clarifying \textit{their} (not your)
        tactical lines of action.
        \item \textbf{Resolution of Judge/Boss:} Credible robust action
        works because the center (Cosimo) appears to have
        \textbf{no unequivocal self-interests}; he `merely' responds to
        requests. Control is diffused, as others infer and serve his
        inscrutable interests.
        \item \textbf{Preconditions:} Requires specific network structures
        for channeling requests and opaque, coherent interests to be credible.
    \end{itemize}

    \subsubsection{Historical Context: Florence (1400-1434)}

    \noindent \textbf{Transition:} From late medieval urban factionalism
to a regionally consolidated Renaissance state.\\

    \noindent \textbf{Ultimate Causes:} Unsuccessful class revolt (Ciompi
revolt, 1378-82) and severe fiscal crisis due to wars (Milan and Lucca wars,
1424-33).\\

    \noindent \textbf{Medici's Rise:} Cosimo de Medici (1389-1464) founded a
dynasty, consolidated a Europe-wide banking network, and sponsored the
Renaissance.\\

    \noindent \textbf{Cosmio's `Sphinx-Like' Character:}\
    \begin{itemize}
        \item Contemporaries deeply appreciated his power, yet eyewitness
        accounts describe him as indecipherable.
        \item He remained in the background, acting through deputies, with
        little known of his direct responsibilities.
        \item He never assumed lasting public office and rarely gave public
        speeches.
        \item His actions appeared reactive, serving his `extremely multiple
        interests'.
        \item His replies were often brief and obscure, sometimes Delphic or
        using proverbs, allowing double interpretations.
    \end{itemize}

    \subsubsection{Methodology and Data}

    \noindent \textbf{Approach:} An `archaeological dig' into the structural
preconditions for Medici's success, focusing on the composition and social
network structure of the Medici party versus their opponents, the `oligarchs'.\\

    \noindent \textbf{Sources:} Based on extensive historical work,
especially Dale Kent's `The Rise of the Medici' (1978). Supplemented by
1427/1433 catasto (tax registers), Najemy (1982), and Martines (1963).\\

    \noindent \textbf{Data Types (9 types of relations among elite families):}
    \begin{itemize}
        \item \textbf{Kinship:} Intermarriage ties (assymetric, 1394-1434).
        \item \textbf{Economic:} Trading/business ties, joint ownerships,
        bank employment, real estate ties.
        \item \textbf{Political:} Patronage and personal loans (multifaceted
        motives).
        \item \textbf{Personal:} Friendship, mallevadori (surety) ties.
    \end{itemize}

    \noindent \textbf{Definitions:}
    \begin{itemize}
        \item \textbf{Family:} Operationalized as `people with a common last
        name' (clan), consistent with Florentine social reality and data
        limitations.
        \item \textbf{Elite:} Families meeting any of three criteria:
        \begin{itemize}
            \item 2+ members speaking in Consulte e Prtiche (1429-34);
            \item 3+ members qualified for scrutiny (election to high
            office) in 1433;
            \item Magnate clan (215 families identified, 92 for network
            analysis).
        \end{itemize}
        \item \textbf{Blockmodel Analysis:} Aggregates actors (families)
        into structurally equivalent `blocks' based on common external ties
        with outsiders, not dense internal relations (cliques). Used to
        visualize marriage and economic networks (`strong ties', fig. 2a)
        and political/friendship networks (`weak ties', fig 2b).
    \end{itemize}

    \subsubsection{Empirical Findings: Attributional vs. Network Structure}

    \noindent \textbf{Attributional Analyses (Challenging Traditional Views):}
    \begin{itemize}
        \item \textbf{Economic Class (Wealth/Change in Wealth):} Both
        Mediceans and oligarchs were wealthy, but their wealth distributions
        were statistically identical and highly heterogenous. Not a Marxist
        class struggle.
        \item \textbf{Social Class (Presitge/Political Age):} Oligarch were
        more skewed towards older participants due to the \textit{absence}
        of `new men' from their party, not absence of particians from the
        Medicean side. Mediceans were more socially heterogenous, and
        relative to neutrals, were distinctly `old-guard patrician'.
        \item \textbf{Neighbourhood Residence:} No statistically significant
        difference, both parties mirrored each other in geographical
        concentration, especially in San Giovanni. San Giovanni was the most
        polarized quarter.
        \item \textbf{Conclusion:} There was a \textbf{structural mismatch}
        between contemporaries' clear cognitives typifications (oligarchs as
        old, wealth patricians; Mediceans as heroes of rising new men) and
        the objective heterogeneity and overlap of social groups at the
        behavioural level. Classical group theories of parties (
        pluralist/neo-Marxist) are insufficient.
    \end{itemize}

    \noindent \textbf{Social Network Structure (Blockmodel Analysis):}
    \begin{itemize}
        \item \textbf{Predictive Power:} Marriage and economic blockmodels
        remarkably predict political partisanship, despite attributional
        identity. The Medici family itself bridged both sides.
        \begin{itemize}
            \item 93\% of families within the `Medicean circle' were Medici
            partisans.
            \item 82\% of other partisan families (excluding neutrals)
            joined the oligarch side.
        \end{itemize}
        \item \textbf{Medici Party Structure:} An extraordinarily
        \textbf{centralized `star' or `spoke' network system}.
        \begin{itemize}
            \item Medici partisans were connected to other partisans and the
            oligarch elite almost solely \textit{through the Medici}.
            \item Medici partisans had remarkably few intraelite network
            ties, being `structurally impoverished'.
        \end{itemize}
        \item \textbf{Oligarch Party Structure:} Densely interconnected,
        especially through marriage, but this density led to
        \textbf{cacophony and cross-pressure}, not cohesive collective
        action (e.g., Rinaldo Albizzi's failed mobilization).
        \item \textbf{The Structural Atomization Puzzle:} Why did this
        centralized spoke system maintain itself.
        \begin{itemize}
            \item Medici followers had clear incentives to form cross-ties
            to alleviate dependence.
            \item The Medici discouraged
            \textbf{multiplex ties} (overlapping marriage and economic relations) with their followers, and also segregation of types of ties with the Medici themselves.
        \end{itemize}
        \item \textbf{Resolution: Double Segregation of Attributes}
        \begin{itemize}
            \item \textbf{Patrician Supporters:} Wealthy, old patricians (
            e.g., Guicciardini, Tornabuoni blocks) intermarried with the
            Medici, but resided \textit{outside} the San Giovanni quarter.
            \item \textbf{New Men Supporters:} Connected to Medici through
            economic or personal loan ties (e.g., Ginori, Orlandini,
            Cocco-Donati blocks), but residede \textit{within} San Giovanni.
            \item \textbf{Mechanism:} Patricians and new men despised each
            other and had limited interaction. Only the Medici linked these
            segments. This structural isolations inhibited defensive
            counter-alliances (`revolt of the colonels').
            \item \textbf{Medici Strategy:}
            \begin{itemize}
                \item In marriage and friendship, Medici were highly
                selective (snobbish, marrying other patricians).
                \item In the economic sphere, they associated heavily with
                new men, unlike other elite families.
                \item Their distinctiveness was
                \textit{associating with enw men at all}, not representing them.
            \end{itemize}
            \item \textbf{New Men's Responsiveness:} 90-96\% of new men
            economically/politically tied to Medici became active partisans.
            This was not due to active Medici mobilization of new men as a
            whole, but the \textit{oligarchs extraordinary inaction} towards
            them, leaving new men `structurally available'. Oligarchs'
            polemics branded Medici as `class traitors' (``heroes of then
            new men'').
        \end{itemize}
        \item \textbf{Control Mechanisms:}
        \begin{itemize}
            \item Spoke structure ensured dependence and channeled
            communication through Medici.
            \item Double segregation prevented counter-alliances among
            partisans.
            \item Formal affine relations with distant patricians (less
            frequent contact) contrasted with friendly, useful business ties
            with local new men (where status gap ensured deference).
            \item Attributional heterogeneity made Medici a potent `swing vote'.
            \item \textbf{Contradiction was key to control}, especially with
            intense surrounding cognitive group identities.
        \end{itemize}
    \end{itemize}

    \subsubsection{Network Dynamics: How the Medici Party Emerged}

    \noindent \textbf{No Grand Design:} Cosimo did not design his party or
initially intend to take over the state. The network patterns emerged from
oligarchs' previous actions and inadvertently channeled material to the Medici.

    \begin{itemize}
        \item [$1$.] \textbf{Dynamics of Patrician Marriage (1385-1420):}
        \begin{itemize}
            \item \textbf{Context:} Oligarchs' reconsolidation of control
            after the Ciompi revolt (1378).
            \item \textbf{Historical Trend:} Increasing rates of
            \textbf{neighbourhood exogamy} in Florentine elite marriage.
            This dissolved older `neighbourhood solidarity' mode of elite
            organization (quasi-feudal, intra-neighbourhood hierarchies).
            \item \textbf{Elite Closure (Post-Ciompi, 1382):}
            \begin{itemize}
                \item \textbf{Shunning `Class Traitors':} Patrician
                families (like Medici) who sympathized with Ciompi were
                severely ostracized in marriage by victorious oligarchs.
                This created the \textbf{structural barrier} seen in Figure
                2a between oligarchs and Medicean patrician blocks (
                Guicciardini, Tornabuoni).
                \item
                \textbf{Oligarch Co-optation and Cross-Neighbourhood Cycles:} Oligarchs began forming \textbf{cross-neighbourhood marraige cycles} to co-opt potentially bridging ``swing vote'' families (e.g., Rondinelli). This created a dense, citywide elite, closing in on itself.
                \item \textbf{Outcast Patrician Exogamy:} Structurally
                isolated, outcast patricians (like Medici) were forced to
                marry fellow isolates \textit{outside} their neighbourhoods
                to preserve status, leading to higher exogamy rates among them.
            \end{itemize}
            \item \textbf{Medici's Anomalous Position:}
            \begin{itemize}
                \item \textbf{Survival:} Veieri di Cambio's (Medici clan
                head) defusion of a pro-Alberti revolt in 1393 and Giovanni
                de Medici's later circumspection (avoiding politics,
                squelching discontent) saved the Medici name from utter
                ostracism. This earned them begrudging oligarch gratitude.
                \item \textbf{Limited Co-optation:} Oligarchs slowly
                relented in the 1420s, allowing some intermarriage with
                Medici (e.g., Albizzi, Gianfigliazzi blocks), but only after
                the Medici were already deeply isolated.
                \item \textbf{Exploiting Structural Holes:} Oligarchs
                over-focused on containing San Giovanni (Ricci's old home),
                creating a `structural hole' in the Santo Spirito quarter.
                The Medici gradually exploited this, directing 100\% of
                their own marriages to Santo Spirito by the early 1430s,
                often `wife-receiving' (less status-picky).
            \end{itemize}
            \item \textbf{Adaptive Learning:} Elite tactics evolved not from
            grand strategies but as a mutually adaptive learning process,
            with families making `boundedly rational local action' from
            their egocentric network positions.
        \end{itemize}
        \item [$2$.] \textbf{Dynamics of New Men Economic Ties (1420s-1430s):}
        \begin{itemize}
            \item \textbf{Catalyst:} Milan and Lucca wars (1424-33) led to
            devastating tax extraction, threatening family patrimonies.
            \item \textbf{Neighbourhood Politics Revival:} Tax assessments
            by neighbourhood intensified local politics.
            \item \textbf{Oligarch Repression:} Patricians legislatively
            targeted new men; successful repression abolished new men's
            nascent corporate forms (religious confraternities), leaving
            them without local support.
            \item \textbf{Medici as Exception:} Locked in by their dense
            marriage network, most oligarchs rejected appeals from new men.
            The Medici, with their `structurally contradictory position',
            had the discretion to respond to pleas from San Giovanni new men.
            \item \textbf{Medici Self-Consciousness:} The surge of
            supplication from San Giovanni new men during the Milan war
            galvanized the Medici into self-awareness as a political party.
            Oligarch actions (e.g., Rinaldo Albizzi's class alliance request
            to Giovanni de Medici, which Giovanni equivocally refused)
            further solidified the Medici as a distinct faction.
        \end{itemize}
    \end{itemize}

    \subsubsection{Network Identities: Robust Action and Legitimacy}

    \noindent \textbf{Credibility of Robust Action:} The contradictory
attributions of `Medici self-interest' (hero of new men vs. patrician) were
credible because the disparate groups of Medici supporters rarely hd the
opportunity to compare notes privately. Even if they had, low trust would
have prevented agreement.\\

    \noindent \textbf{Opaque Self-Interests:} Medici goals (money, prestige,
power) were tied to specific roles, not an overarching utility function. In
chaotic times, the ``games themselves are all up for grabs'', making
revealed preferences impossible to infer \textit{a priori}. Cosimo's and
Giovanni's ``shrewed and multivocal opportunism'' was a feature of varying
game structures, not fixed personal goals.\\

    \noindent \textbf{Legitimacy (Pater Patriae):}
    \begin{itemize}
        \item Cosimo was enshrined as \textit{pater patriae} upon his death,
        transmuting his ambiguity into public interest.
        \item He achieved this not by directly defeating oligarchs or
        slinging mud, but by \textbf{positional maneuvering}.
        \item His `reactive character' forced oligarchs into aggressive,
        self-interested offensive actions (e.g., repression, attempted
        seizure of city hall).
        \item During the fiscal crisis, Cosimo channeled his bank's assets
        into state debt funding, gaining the appearance of Florence's
        `financial saviour' while his opponents faced ruin.
        \item This led new men and political neutrals to delegitimize the
        oligarchs (labeling them as self-interested) and embrace Cosimo as
        saviour of the republic, leading to his triumphal recall from exile.
        \item His robust, multivocal actions gained him the ``legitimizing
        aureole of protector of the status quo'', transforming his party
        into a state.
    \end{itemize}

    \subsubsection{Conclusion}

    \noindent State centralization adn the Renaissance arose form tumultuous
historical events filtered through elite transformation. Cosimo did not
create the Medici party but shrewdly learned network rules. He used a shroud
of \textbf{multiple, impenetrable identities} to maintain robust discretion
and Solomonic legitimacy.\\

    \noindent Understanding state building requires delving into the
\textbf{relational substratum} of lives, recognizing the
\textbf{localized, ambiguous and contradictory nature} of actions, networks,
and identities. This heterogeneity explains the birth of political power.


    \section{Communities}

    \subsection{Marina G Duque. Recognizing international status: A relational approach. Inter-national Studies Quarterly, 62(3):577–592, 2018.}

    \subsubsection{Central Argument and Thesis}
    \noindent \textbf{Problem:} International relations scholarship relies
on the concept of `status' to explain phenomena like war and foreign policy,
but lacks a clear understanding of what status is and how it is achieved.\\

    \noindent \textbf{Critique of Conventional View:} Previous research
treats status as a function of a state's attributes, particularly material
capabilities like wealth and military power. This approach is a form of
material reductionism and fails to capture the social nature of status.\\

    \noindent \textbf{Author's Thesis (Relational Approach):} Status is not
derived from state attributes but from \textbf{social recognition}. It is a
relational process where a state gains admission into a `club' after being
deemed to follow it rules of membership. Status is therefore influenced by
two key social processes:
    \begin{itemize}
        \item \textbf{Self-reinforcing Dynamics:} Recognition breeds more
        recognition.
        \item \textbf{Social closure:} A state's existing relationships and
        its similarity to other states influence its ability to achieve status.
    \end{itemize}

    \subsubsection{The Conventional (Attribute-Based) Approach and Its Flaws}

    \noindent \textbf{Definition:} Defines status as a state's ranking on
valued attributes, such as economic military, and technological capabilities.\\

    \noindent \textbf{Key Flaws:}
    \begin{itemize}
        \item \textbf{Material Reductionism:} Reduces status to material
        power, making the concept analytically redundant.
        \item \textbf{Fetishism:} Mistakenly treats social relations as
        inherent properties of states. It equates status with possessing
        symbols (e.g., nuclear weapons), but these attributes have no
        intrinsic value without social agreement.
        \item \textbf{Reification:} Treats the status order as external to
        states, making status achievement an autonomous act rather than a
        social process of recognition.
        \item \textbf{Empirical Mismatch:} Fails to explain why some states
        with significant material resources receive low status (e.g., North
        Korea as a `rogue state') or why others receive more recognition
        than their capabilities would suggest (e.g., Italy, Egypt).
    \end{itemize}

    \subsubsection{A Relational Theory of Status}

    \noindent \textbf{Definition (from Max Weber):} Status is an ``effective
claim to social esteem in terms of positive or negative privileges''. This
has four key dimensions:
    \begin{itemize}
        \item [$1$.] \textbf{Effective Claim (Recognition):} A state's claim
        to status must be recognized by others to be effective. Status
        involves admission into a `club' based on following its distinctive
        lifestyle and rules.
        \item[$2$.] \textbf{Social Esteem (Symbolic):} Status is based on
        `social honour', which can be attached to any symbol, material or
        ideational. The value of attributes is socially constructed, not
        intrinsic.
        \item[$3$.] \textbf{Social Hierarchy (Privileges):} The status order
        is a hierarchy that grants privileges to high-status members and
        disadvantages to low-status ones. This is maintained through
        \textbf{social closure}, where high-status groups form dense
        internal ties and justify their privileges based on their
        distinctiveness.
        \item[$4$.] \textbf{Conventions (Practices):} The status order is
        regulated by conventions (norms and rules) that emerge from the
        practices of states themselves.
    \end{itemize}

    \noindent \textbf{Two Key Relational Processes:}
    \begin{itemize}
        \item [$1$.] \textbf{Status is Self-Reinforcing:} Because status
        comes from peer attribution, states that are already highly
        recognized are more likely to attract additional recognition. This
        is a purely structural effect.
        \item[$2$.] \textbf{Social Closure Shapes Status:} This has two
        implications:
        \begin{itemize}
            \item \textbf{Connectedness matters:} A state's existing
            relationships, especially with high-status actors, influence its
            ability to gain status.
            \item \textbf{Similarity begets recognition:} States tend to
            recognize other states that have similar values (e.g.,
            democracy) and resources, a concept known as \textbf{homophily}.
        \end{itemize}
    \end{itemize}

    \subsubsection{Empirical Strategy and Methodology}

    \noindent \textbf{Data:} The network of diplomatic embassies from the
Diplomatic Contacts Database.
    \begin{itemize}
        \item \textbf{Rationale:} Establishing an embassy is a costly and
        symbolic act of recognition. The network of embassies provides a
        comprehensive map of recognition practices.
    \end{itemize}

    \noindent \textbf{Unit of Analysis:} The
\textbf{network of embassies} (a relational measure), not just a state-level
count of embassies received. This preserves information about the structure
of relationships (i.e., who recognizes whom).\\

    \noindent \textbf{Method: Temporal Exponential Random Graph Model (TERGM):}
    \begin{itemize}
        \item \textbf{Advantage:} This model can test for both
        attribute-based (exogenous) effects and relational/structural (
        endogenous) effects, such as reciprocity and popularity. It also
        avoids statistical biases common when analyzing relational data.
    \end{itemize}

    \subsubsection{Key Findings}
    \noindent \textbf{The Relational model performs much better} in
explaining embassy ties than conventional attribute-based models.\\

    \noindent
\textbf{Relational dynamics are powerful drivers of status recognition:}
    \begin{itemize}
        \item \textbf{Popularity (Self-Reinforcement):} States with more
        embassies are significantly more likely to attract additional ones.
        A state that is one standard deviation above the mean in embassies
        received is \textbf{2.4 times more likely} to get another one.
        \item \textbf{Reciprocity:} States are
        \textbf{almost six times more likely} to have an embassy in another
        country if that country reciprocates.
        \item \textbf{Transitivity:} States are more likely to establish an
        embassy where their existing diplomatic partners also have a presence.
    \end{itemize}

    \noindent \textbf{Similarity (Homophily) is crucial:} States are
significantly more likely to recognize other states that are simialr to them
in terms of democracy, human rights, economic freedom, and military
spending. The relevancce of attributes is socially defined.\\

    \noindent
\textbf{State attributes have mixed and often counter-intuitive effects:}
    \begin{itemize}
        \item While military capability positively correlates with
        recognition, higher wealth (GDP per capita) and possessing nuclear
        weapons are associated with a \textit{reduced} likelihood of
        receiving an embassy.
    \end{itemize}

    \noindent \textbf{Fundamental Values Matter:} Values like democracy,
human rights, and economic liberalism are at least as important for gaining
recognition as material resources.

    \subsection{Wayne W Zachary. An information flow model for conflict and fission in small groups. Journal of Anthropological Research, 33(4):452–473, 1977.}

    \subsubsection{The Problem of Fission in Small Groups}

    \noindent \textbf{Central Issue:} How and why fission (splitting) occurs
in small, bounded groups is a long-standing topic of study in social
anthropology.\\

    \noindent \textbf{Traditional Perspectives:} Fission has been studied
through lenses like descent theory and ecological adaptation. \\

    \noindent \textbf{Proposed Model:} This study uses a new formal model
based on \textbf{information flow} within a social network to analyze and
predict fission.\\

    \noindent \textbf{Core Thesis:} Fission is the result of an
\textbf{unequal flow of sentiments and information} across the ties in a
social network. This differential sharing leads to the formation of
subgroups with greater internal stability, eventually causing a split.

    \subsubsection{Case Study: A University Karate Club}

    \noindent \textbf{Context:} The model was developed using data from a
university karate club observed over three years (1970-1972).\\

    \noindent \textbf{The Conflict:} A dispute arose between the club's
instructor, \textbf{Mr. Hi}, and the president, \textbf{John A.}, over the
price of lessons.\\

    \noindent \textbf{Factionalization:} The club divided into two
ideological factions:
    \begin{itemize}
        \item \textbf{Mr. Hi's supporters:} Viewed him as a spiritual mentor.
        \item \textbf{John A.'s supporters:} Viewed Mr. Hi as a paid
        employee attempting to raise his salary.
    \end{itemize}

    \noindent \textbf{Process of Fission:}
    \begin{itemize}
        \item Factions were not formally organized but emerged from existing
        friendship networks during political crises.
        \item Crises strengthened friendship bonds \textit{within} factions
        and weakened bonds \textit{between} them through selective interaction.
        \item After a series of confrontations, the officers, led John A.,
        fire Mr. Hi.
        \item Mr. Hi's supporters resigned and formed a new club, completing
        the fission.
    \end{itemize}

    \subsubsection{Social Network Models and Their Limitations}

    \noindent \textbf{Standard Social Network Model:} Represents individuals
as points (\textbf{nodes}) and their relationships as lines (\textbf{edges}).
    \begin{itemize}
        \item \textbf{Graph Representation:} A visual diagram of nodes and
        edges.
        \item \textbf{Matrix Representation:} A square matrix of 0s and 1s,
        where a 1 indicates an existing relationship between the individuals
        presented by the row and column.
    \end{itemize}

    \noindent \textbf{Key Limitation:} Standard social network analysis is \textbf{static} and structural, ignoring the \textbf{processual aspects} of social activity. It is therefore inadequate for analyzing the process of fission.

    \subsubsection{The Capacitated Network Model}

    \noindent \textbf{A Processual Model:} This model views the network as a
system of channels for information flow, enabling a processual analysis.\\

    \noindent \textbf{Model Specification:} Defined as an ordered triplet \textbf{(V, E, C)}.
    \begin{itemize}
        \item \textbf{V:} The set of individuals (nodes) in the network;
        \item \textbf{E:} The \textbf{existence matrix}, a standard 0/1
        social network matrix indicating if a tie exists.
        \item \textbf{C:} The \textbf{capacity matrix}, which quantifies the
        relative strength (or potential for information flow) of each
        existing edge.
    \end{itemize}

    \noindent \textbf{Determining Edge Capacity:}
    \begin{itemize}
        \item Directly measuring ``political information'' flow is impossible.
        \item \textbf{Assumption:} The amount of information transmitted
        between two individuals is a \textbf{linear function} of the number
        of different social contexts in which they interact.
        \item \textbf{Method:} A scale of 8 interaction contexts was
        created (e.g., academic classes, bars, karate tournaments). The
        value for each edge in matrix C is the total number of contexts in
        which that pair of individuals was observed interacting.
    \end{itemize}

    \subsubsection{Analysis and Results}

    \noindent \textbf{Technique:} The \textbf{maximum flow-minimum cut}
labeling procedure.
    \begin{itemize}
        \item This technique models information as flowing from a
        \textbf{source} (Mr. Hi) to a \textbf{sink} (John A.).
        \item A \textbf{a minimum cut} is the `bottleneck' in the network --
        the break with the smallest capacity that separates the source from
        the sink.
        \item The \textbf{Max-Flow Min-Cut Theorem} states that the maximum
        possible flow between a source and a sink is equal to the capacity
        of the minimum cut.
    \end{itemize}

    \noindent \textbf{Hypotheses Tested:}
    \begin{itemize}
        \item [1.] The minimum cut in the network will align with the
        factional divide.
        \item[2.] The minimum cut will accurately predict membership in the
        two new clubs formed after the fission.
    \end{itemize}

    \noindent \textbf{Findings:} The model was highly accurate.
    \begin{itemize}
        \item It predicted faction membership with \textbf{100\% accuracy}.
        \item It predicted membership in the post-fission clubs with \textbf{
        97\% accuracy}.
        \item The single error involved an individual who was structurally
        part of John A.'s faction (and correctly placed on the sink side fo
        the cut by the model) but joined Mr. Hi's club for an overriding
        personal reason (an upcoming black belt test).
    \end{itemize}

    \noindent \textbf{Uniqueness:} The minimum cut was proven to be unique
by reversing the network (source becomes sink) and re-running the analysis,
which identified the same cut.

    \subsubsection{Conclusion and Implications}

    \noindent \textbf{The Mechanism of Fission:} The model reveals a
`vicious cycle' feedback mechanism.
    \begin{itemize}
        \item \textbf{Organizational Feedback:} Factional behaviour
        strengthens the network's structural `cut' or bottleneck.
        \item \textbf{Ideological Feedback:} The network bottleneck prevents
        the flow of information and shared understanding between factions,
        which in turn intensifies ideological division.
    \end{itemize}

    \noindent \textbf{Generalizability of the Model:}
    \begin{itemize}
        \item The presence of a \textbf{unique minimum cut} in a capacitated
        network model of any small group may indicate a structural weakness
        that could lead to fission.
        \item The model is flexible and can be built using different types
        of relationships (e.g., kinship) and other methods for assigning
        edge capacity.
    \end{itemize}

    \noindent \textbf{Power of the Model:} The capacitated network model is
more powerful than standard network models because it is \textbf{processual}, not static, and incorporates more information (the capacity matrix C). It views the social system as an information processor.

    \section{Communities II}

    \subsection{Roger V Gould. Multiple networks and mobilization in the Paris Commune, 1871.
American Sociological Review, pages 716–729, 1991.}

    \subsubsection{Core Argument and Critique of Existing Research}

    \noindent \textbf{Thesis:} Successful social movement mobilization
depends not on the sheer number of social ties, but on the
\textbf{interplay between formal organizational networks and pre-existing informal social networks}. The \textbf{structure} of these networks is a crucial factor.\\

    \noindent \textbf{Critique of Prior Social Movement Research:}
    \begin{itemize}
        \item Research often treats network factors as individual-level
        variables by simply counting a person's social ties and using that
        count in regression models.
        \item This reductionist approach obscures two critical issues:
        \begin{itemize}
            \item [$1$.] \textbf{Network Structure:} The way social
            relations are patterned, which can create effects on a
            supra-individual level.
            \item[$2$.] \textbf{Network Multiplexity:} The existence of
            multiple, overlapping types of social ties (e.g., organizational
            and informal) between individuals.
        \end{itemize}
        \item By focusing on a single network, previous studies have often
        neglected how pre-existing informal ties continue to matter even
        late in the mobilization process.
    \end{itemize}

    \subsubsection{Case Study: The Paris Commune (1871)}

    \noindent \textbf{Historical Context:}
    \begin{itemize}
        \item Emerged from social discontent and France's defeat in the Franco-Prussian War.
        \item A conflict arose between the French national government under
        Adolphe Thiers and the people of Paris over the artillery of the \textbf{Paris National Guard}, a popular militia.
        \item The Commune, proclaimed on March 26, 1871, was a two-month
        experiment in democratic socialism that instituted policies like
        workers' cooperatives and universal free education.
        \item It was brutally suppressed during the
        \textbf{`semaine sanglante'} (bloody week) of May 21-28, resulting
        in approximately 15,000 deaths among Parisians.
    \end{itemize}

    \noindent \textbf{Relevance as a Case Study:}
    \begin{itemize}
        \item Insurgent mobilization occurred through a visible formal
        organization, the \textbf{Paris National Guard}.
        \item The Guard's structure allows for a critical comparision:
        \begin{itemize}
            \item Most battalions were recruited along neighbourhood lines (\textbf{residential battalions}).
            \item An important exception was 35 \textbf{volunteer battalions}, which recruited members without regard to their place of residence.
        \end{itemize}
        \item This setup permits a direct examination of how the
        \textbf{organizational network} (the Guard) interacted with the \textbf{informal network} (neighbourhood ties).
    \end{itemize}

    \subsubsection{Analysis: The Interaction of Formal and Informal Networks}

    \noindent \textbf{The Role of Informal Neighbourhood Ties:}
    \begin{itemize}
        \item The residential recruitment policy was crucial because it
        linked the formal organization of the Guard to the informal social
        networks of neighbourhoods.
        \item Insurgents expressed a powerful, neighbourhood-based identity,
        with some units refusing to fight outside their home \textit{quartier}.
        \item While historians have often viewed this `localism' as a
        strategic military weakness, it was a fundamental
        \textbf{source of solidarity and commitment} to the insurgent cause.
    \end{itemize}

    \noindent
\textbf{Evidence from Arrest Patterns (Residential vs. Volunteer Battalions):}
    \begin{itemize}
        \item \textbf{Finding:} During the \textit{semaine sanglante},
        residential battalions demonstrated greater cohesion and sustained
        resistance for longer than volunteer battalions.
        \item \textbf{Data Analysis:} A comparison of cumulative arrests for
        the two battalion types during the week of fighting.
        \begin{itemize}
            \item \textbf{Early Surrender:} By second day, 45.2\% of
            arrested volunteers had been detained, compared to only 26.4\%
            for residential units. This suggests volunteer units surrendered
            more readily in the initial, less intense phase of fighting.
            \item \textbf{Sustained Resistance:} In the final two days, 42.7\% of total arrests for residential units occurred, compared to 31.9\% for volunteer units, indicating that neighbourhood-based units fought on later into the week.
        \end{itemize}
        \item \textbf{Conclusion:} Pre-existing informal ties were critical
        for maintaining solidarity even in the final moments of the
        insurrection, demonstrating that they are not just important for
        initial recruitment.
    \end{itemize}

    \subsubsection{Network Autocorrelation Model}

    \noindent \textbf{Objective:} To quantitatively test if levels of
resistance in different neighbourhoods were interdependent due to the
network of organizational ties.\\

    \noindent \textbf{Methodology:}
    \begin{itemize}
        \item \textbf{Model:} A network autocorrelation regression model
        $(y = \rho Wy + X\beta + \epsilon)$ was used. This model posits that
        the outcome in one district $(y)$ is influenced by exogenous
        variables $(X)$ and by the outcomes in other districts to which it
        is linked $(Wy)$.
        \item \textbf{Dependent Varaibles $(y)$:} Two measures of resistance
        were used: (1) average battalion size and (2) deaths per 1,000
        inhabitants.
        \item \textbf{Key Network Matrix $(W)$:} A matrix representing \textbf{enlistment overlaps} -- the number of Guardsmen from one district who served in the legion of another district.
        \item \textbf{Comparison:} The results of the enlistment network
        model were compared to a spatial model that used a simple
        geographical adjacency matrix to check if effects were merely due to
        proximity.
    \end{itemize}

    \noindent \textbf{Key Findings:}
    \begin{itemize}
        \item
        \textbf{A positive and significant autocorrelation effect $(\rho)$ was found for the enlistment network model}, but not for the spatial model.
        \item \textbf{Interpretation:} High levels of commitment in one
        district enhanced commitment in other districts to which it was
        connected through enlistment overlaps. This interdependence was
        fundamentally social, not merely geographical.
    \end{itemize}

    \noindent \textbf{The Directionality of Influence:}
    \begin{itemize}
        \item The model was re-estimated using the
        \textbf{transpose of the $W$ matrix $(W')$}, which tests for the
        reverse causal pathway (i.e., whether a district was influenced by
        outsiders serving in its units).
        \item \textbf{Results:} The autocorrelations coefficient became
        insignificant.
        \item \textbf{Conclusion:} The influence was \textbf{unidirectional}
        . A district's resistance level was affected by the experiences of
        its \textbf{own residents serving in other districts}, but not by
        the presence of insurgents from other districts serving within its
        units. Insurgents serving away from home influenced their neighbours
        back home, but not the men they served alongside from other
        neighbourhoods.
    \end{itemize}

    \subsubsection{Theoretical Implications}

    \noindent \textbf{Interaction is Key:} The effectiveness of formal
organizational links is dependent on their
\textbf{interaction with informal social networks}. The cross-neighbourhood
solidarity created by enlistment overlaps only emerged because the
individuals forming those links were also deeply embedded in their own
neighbourhood ties.\\

    \noindent \textbf{Structure and Multiplexity are Inseparable:}
    \begin{itemize}
        \item To understand the effect of enlistment overlaps, one must
        first recognize the structure of both informal groups (
        neighbourhoods) and formal groups (Guard units).
        \item The influence process operated through the entire network
        structure, including direct and indirect links, which would be
        obscured by simply counting ties.
    \end{itemize}

    \noindent \textbf{Generalizing the Findings for Social Movement Theory:}
    \begin{itemize}
        \item Mobilization can be viewed as a `mapping' of formal
        organizational ties onto an existing informal social structure.
        \item
        \textbf{Completely cross-cutting organizations} (like the volunteer battalions) that ignore pre-existing networks struggle to sustain commitment.
        \item \textbf{The intermediate case}, where mobilization largely
        follows informal structures but also creates links across them (as
        in the Paris Commune), provides the greatest potential for formal
        and informal networks to jointly create solidarity
    \end{itemize}

    \subsection{Cesi Cruz, Julien Labonne, and Pablo Querubin. Social network structures and the politics of public goods provision: evidence from the philippines. American Political Science Review, 114(2):486–501, 2020.}

    \subsubsection{Core Research Question and Thesis}

    \noindent \textbf{Research Question:} What is the relationship between a
village's social structure and the political incentives for providing public
goods?\\

    \noindent \textbf{Main Argument (Thesis):} In contexts where
\textbf{politicians, not communities, are responsible for public goods provision}, greater \textbf{social fractionalization can increase public goods provision and electoral competition}.\\

    \noindent \textbf{Mechanism:} Social cohesion (low fractionalization)
can facilitate \textbf{elite capture}. Concentrated social networks give a
few clan leaders strong bargaining power to demand private, clientelistic
transfers from politicians in exchange for votes, which crowds out spending
on public goods. In fragmented villages, this strategy is less effective,
making broad public goods provision a more appealing electoral strategy for
politicians.

    \subsubsection{Context: Local Politics in the Philippines}

    \noindent \textbf{Public Goods Provision:}
    \begin{itemize}
        \item \textbf{Municipal mayors} are primarily responsible for
        delivering local public goods like health centers, schools, and
        infrastructure.
        \item Funding comes mostly from \textbf{central government transfers} (Internal Revenue Allotment), not local taxes. This makes community-level collective action for funding less relevant.
        \item Despite available funds, there is significant under-provision
        of public goods, often attributed to
        \textbf{clientelism} (exchanging private goods/money for votes).
    \end{itemize}

    \noindent \textbf{Social and Political Structure:}
    \begin{itemize}
        \item \textbf{Clans and extended families} are the key actors in
        local politics.
        \item Political alliances require securing the support of clan
        leaders, who leverage norms of reciprocity (e.g., ``debt of
        gratitude'') to deliver blocs of votes.
        \item Political parties are weak; politics is highly personalistic
        and organized around families.
    \end{itemize}

    \subsubsection{Methodology: Identifying Clans and Measuring Fractionalization}

    \noindent \textbf{Data:}
    \begin{itemize}
        \item \textbf{Networks:} Constructed from a survey of
        \textbf{20 million individuals} in over 15,000 villages.
        \item \textbf{Public Goods:} 2010 population census data on the
        presence of schools, markets, health centers, etc.
        \item \textbf{Elections:} Official results for municipal and village
        elections.
    \end{itemize}

    \noindent \textbf{Constructing Family Networks:}
    \begin{itemize}
        \item A \textbf{network tie (edge)} between two families is
        identified when a marriage between members of those families is
        observed.
        \item This is possible due to Philippine naming conventions, where
        individuals carry two family names (father's and mother's maiden
        name), revealing a marriage tie in their own or their parents'
        generation.
    \end{itemize}

    \noindent \textbf{Identifying Clans via Community Detection:}
    \begin{itemize}
        \item Clans are proxied by
        \textbf{``communities'' in the family marriage network} -- groups of
        familieis with dense internal marriage ties and sparse ties to other
        groups.
        \item The \textbf{Girvan-Newman algorithm} is used to identify these
        communities by sequentially removing network edges with the highest
        `betweenness centrality'. The \textbf{Walktrap algorithm} is used as
        a robustness check.
    \end{itemize}

    \noindent \textbf{Measuring Social Fractionalization (SF):}
    \begin{itemize}
        \item The primary independent variable is a
        \textbf{Herfindahl index of social fractionalization}.
        \item \textbf{Formula:} $SF = 1 - \sum^{C}_{c=1}s^2_c$.
        \item \textbf{Interpretation:} The probability that two randomly
        selected families in a village belong to different clans.
    \end{itemize}

    \noindent \textbf{Empirical Strategy:}
    \begin{itemize}
        \item Village level OLS regressions with
        \textbf{municipality fixed effects} to control for all
        municipality-wide factors and focus on variation \textit{within} a
        municipality.
        \item
        \textbf{To address endogeneity} (e.g., public goods attracting new families and alterting networks), an instrumental variables (IV) approach is used. The instrument is a fractionalization measure built from a network of only individuals aged 45+, representing marraiges from a generation prior.
    \end{itemize}

    \subsubsection{Key Findings}

    \noindent \textbf{Fractionalization and Public Goods:}
    \begin{itemize}
        \item There is a
        \textbf{positive and statistically significant correlation between social fractionalization and the provision of public goods}.
        \item A one standard deviation increase in SF leads to a 6\% point
        increase in the probability of a village having a health center (a 10\% increase relative to the mean) and an 8\% point increase for a highschool (a 40\% increase relative to the mean).
        \item The finding is robust to numerous controls (population,
        wealth, etc.) and the IV strategy.
    \end{itemize}

    \noindent \textbf{Fractionalization and Political Competition:}
    \begin{itemize}
        \item
        \textbf{Social fractionalization is positively correlated with greater political competition}.
        \item More fragmented villages have
        \textbf{less concentrated political influence} (i.e., a larger number of different individuals are named by residents as politically influiential).
        \item They also feature \textbf{more candidates} running for village
        office and \textbf{narrower win margins} in both village and mayoral
        elections.
    \end{itemize}

    \noindent \textbf{Ruling Out Alternative Explanations:} The study finds that social fractionalization is
        \textbf{not correlated with preference heterogeneity} or measures of \textbf{collective action}. This supports the theory that elite capture, rather than community disagreement or inaction, is the key mechanism in this context. \\

    \noindent \textbf{Clientelism (Indirect Evidence):} Higher social
fractionalization is associated with \textbf{lower voter turnout}. In the
Philippines, high turnout is strongly correlated with clientelism and
vote-buying, so this suggests
\textbf{less clientelism in more fragmented villages}.




 \end{document}