\documentclass{article}

%settings
\usepackage[utf8]{inputenc}
\usepackage[explicit]{titlesec}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{braket}
\usepackage[margin=1.0in]{geometry}
\usepackage{bbold}
\usepackage{fancyhdr}
\usepackage{fancyvrb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\pagestyle{fancy}
\fancyhead[L]{\leftmark}
\fancyhead[R]{\thepage}

\title{POL325: Contemporary Latin American Politics}
\author{Shamel Bhimani}
\date{Fall 2025}

\begin{document}

\maketitle

\tableofcontents
\newpage

\section{Development and its Alternatives}
    \subsection{Cristóbal Kay (2018) Modernization and Dependency Theory. From The Routledge Handbook of Latin American Development Julie Cupples, Marcel Palomino-Schalscha, and Manuel Prieto, eds. New York: Routledge, pp 15-28.}
    \subsubsection{Modernization Theory}
    \noindent \textbf{Historical Context:} Emerged in the North (
1950s-1960s) during the Cold War. Arose post-WWII as decolonization
accelerated.\\

    \noindent \textbf{Core Idea:} Development is a transition from a
`traditional' to a `modern' society. It posits that underdeveloped countries
can `catch up' by replicating the experience of Western, developed countries.
    \begin{itemize}
        \item Change is seen as determined by internal factors
    \end{itemize}

    \noindent \textbf{Dualistic Typology:}
    \begin{itemize}
        \item \textbf{Traditional Socieites:} Simple, rural subsistence
        economy, family labour, primitive technology, low productivity.
        Characterized by partciularism, adscription, and collective
        orientation. `Traditionalism' itself is seen as a barrier due to
        fatalistic outlook.
        \item \textbf{Modern Societies:} Complex, industrial,
        market-oriented, wage labour, scientific technology, high
        productivity. Characterized by universalism, achievement
        orientation, self-orientation, upward social mobility, and rule of law.
    \end{itemize}

    \noindent \textbf{Key Theorists and Concepts:}
    \begin{itemize}
        \item \textbf{Walt W. Rostow (1960):} Proposed five universal stages
        of economic growth:
        \begin{itemize}
            \item [$1$.] The traditional society;
            \item[$2$.] The preconditions for take-off;
            \item[$3$.] \textbf{Take-off (the key turning point)};
            \item[$4$.] The drive to maturity;
            \item[$5$.] The age of mass-consumption.
        \end{itemize}
        \item \textbf{Samuel Huntington (1968):} Priotized
        \textbf{political order and stability} above other modernization
        goals, concerned that rapid social change could overwhelm political
        institutions. Critiqued mainstream MT for being too static, arguing
        all societies combine traditional and modern elements.
        \item \textbf{Other theorisits focused on:} Value changes (Moore),
        personality transformation (Lerner), psychological factors like the
        desire to achieve (McClelland), and entreprenurial spirit (Hagen).
    \end{itemize}

    \noindent \textbf{Modernization Theory in Latin America:}
    \begin{itemize}
        \item Largely absorbed uncritically by Latin American social
        scientists and policymakers.
        \item \textbf{Gino Germani (1981):} A notable exception who adapted
        MT. He argued that transition processes create conflicts and
        `asynchronies' as different social spehres change at different
        speeds. However, his work was criticized from a Marxist perspective
        for failing to address class and ethnic conflicts.
    \end{itemize}

    \noindent \textbf{Critique of Modernization Theory (by Andre Gunder Frank):}
    \begin{itemize}
        \item Empirically faulty and theoretically weak.
        \item \textbf{Main flaw:} Assumes underdevelopment is an original
        state and ignores how development and underdevelopment are part of a
        single process in the formation of the world capitalist system since
        the 15th century.
        \item Fails to account for the impact of colonialism and imperialism.
    \end{itemize}

    \subsubsection{Dependence Theory (DT)}
    \noindent \textbf{Historical Context:} Arose in Latin America in the
mid-1960s, challenging MT. Influenced by theories of imperialism and the
Latin American structuralist school (ECLAC).\\

    \noindent \textbf{Core Idea:} Underdevelopment is not an original state
but a \textbf{`conditioning situation'} where the economies of some
countries (the periphery) are conditioned by the development and expansion
of others (the center).
    \begin{itemize}
        \item Development and underdevelopment are seen as two faces of the
same historical process of global capitalism.
        \item It analyzes the link between external (global capitalism) and
        internal (class structure, politics) factors.
    \end{itemize}

    \noindent \textbf{Precursor: Raul Prebisch and ECLAC Structuralism:}
    \begin{itemize}
        \item Developed the \textbf{center-periphery paradigm}.
        \item \textbf{Prebisch-Singer Thesis:} Argued the international
        trade system benefits the center at the expense of the periphery due
        to the long-term deterioration of the periphery's terms of trade (
        prices of its primary commodity exports fall relative to the
        industrial goods it imports).
    \end{itemize}

    \noindent \textbf{Main Strands of Dependence Theory:}
    \begin{itemize}
        \item [$1$.] \textbf{Structuralis Strand:} Seeks to reform the
        capitalist system. Uses heterodox development theory concepts.
        \item[$2$.] \textbf{Marxist Strand:} Argues dependency can only be
        overcome by overthrowing capitalism adn transitioning to socialism.
        Relies on historical materialism and the labour theory of value.
    \end{itemize}

    \noindent \textbf{Key Structuralist Theorists and Concepts:}
    \begin{itemize}
        \item \textbf{Osvaldo Sunkel:} Focused on how transnational
        corporations (TNCs) deepen dependece and cause
        \textbf{`national disintegration'}. TNCs weaken the national
        bourgeoisie, fragment society, and shape public policy against the
        national interst.
        \item \textbf{Celso Furtado:} Analyzed
        \textbf{`dependent consumption patterns'}. The consumption habits of
        the rich, influenced by developed countries, create a wasteful,
        capital-intensive, and import-demanding industrial structure that
        perpetuates income concentration and underdevelopment.
        \item \textbf{Fernando Henrique Cardoso and Enzo Faletto:}
        \begin{itemize}
            \item Emphasized analyzing \textbf{`situations of dependency'}
            rather than a single theory, focusing on internal manifestations.
            \item Characterized the process as
            \textbf{`dependent development'}, rejecting stagnationist views
            and acknowledging that economic growth could occur within a
            dependency framework.
            \item Argued a \textbf{`new dependency'} emerged under
            corporatist-authoritarian states controlled by a militarized
            technocratic bureaucracy.
        \end{itemize}
    \end{itemize}

    \noindent \textbf{Key Marxist Theories and Concepts:}
    \begin{itemize}
        \item \textbf{Theotonio Dos Santos:} Identified a
        \textbf{`new character of dependency'} rooted in industrial and
        technological dependence. The lack of a domestic capital goods
        industry and indigenous technological capacity prevents dependent
        economies form being `articulated' and achieving autonomous development.
        \item \textbf{Ruy Mauro Marini:} Focused on \textbf{unequal exchange} (transfer of surplus value to dominate countries) and the resulting \textbf{over-exploitation of labour} in dependent countries to maintain profit rates. Also developed the concept of \textbf{sub-imperialism}, where a larger dependent country like Brazil under its military regime could engage in imperialist practices toward weaker neighbours to solve its own problems of insufficient internal demand.
        \item \textbf{Andre Gunder Frank:} Coined the phrase
        \textbf{``the development of underdevelopment}. Argued that the
        metropolis-satellite linkages constantly recreate underdevelopment.
        Contended that Latin America has been capitalist since the colonial
        conquest, challenging the prevailing feudal/semi-feudal
        characterization and the political strategies based on it. His
        thesis was famously critiqued by Ernesto Laclau for overemphasizing
        market circulation while neglecting relations of production.
    \end{itemize}

    \noindent \textbf{Decline and Legacy of Dependence Theory:} It's
influence waned with the economic crisis of the 1980s and the rise of
neoliberalism. Remebered as the
\textbf{first major challenge to the Eurocentric character of the social sciences} to achieve global influence. It inspired a new generation of scholars to think about development from the perspective of the South.

\subsection{Laura Zapata-Cantu and Fernando González (2021) Challenges for Innovation and
Sustainable Development in Latin America: The Significance of Institutions and
Human Capital. Sustainability.}

    \subsubsection{Innovation and Sustainable Development in Latin America: Core Challenges}

    \noindent \textbf{Context:} Sustainable development is a critical 21st
century challenge, compounded by the COVID-19 pandemic which has shifted
government and business priorities. Innovation is essential for
transitioning to a more sustainable world.\\

    \noindent \textbf{Regional Challenges:}
    \begin{itemize}
        \item \textbf{High social inequality and poverty} remain significant
        obstacles.
        \item The region has a history of
        \textbf{political instability and corruption}, which impacts
        institutional performance in innovation and sustainability.
        \item Economies are vulnerable to \textbf{commodity price volatility} and uncertainty.
        \item There is a \textbf{marginal contribution to global innovation}, often measured by patent registrations.
        \item Latin American countries face ongoing issues with nutrition,
        sanitation, quality education, and economic modernization.
    \end{itemize}

    \subsubsection{Theoretical Framework for Analysis}

    \noindent Two key theoretical perspectives are used to analyze
innovation and sustainable development in the region:
    \begin{itemize}
        \item [$1$.] \textbf{Dynamic Capabilities (DCV):} A framework
        focusing on a
        country's or firm's ability to adapt and transform in response to a
        changing environemnt. It is composed of three core capabilities:
        \begin{itemize}
            \item \textbf{Sensing:} Identifying opportunities and threats by
            scanning markets and technologies. THis aligns with the Global
            Innovation Index (GII) dimensions of \textit{Institutions} and \textit{Market Sophistication}.
            \item \textbf{Seizing:} Mobilizing resources to capture
            opportunities by investing in technology, human capital, and new
            business models. This aligns with the GII dimensions of
            \textit{Human capital and research, Infrastructure, and Business sophistication}.
            \item \textbf{Transforming:} Continuously reconfiguring assets
            and organizational structures to maintain the competitiveness.
            This aligns with the GII dimensions of
            \textit{Knowledge and technology outputs and creative outputs}.
        \end{itemize}
        \item[$2$.] \textbf{Mission-Oriented Policies:} Systemic public
        sector initiatives aimed at solving specific societal problems (
        e.g., climate change) by mobilizing innovation across multiple
        sectors and actors.
        \begin{itemize}
            \item These policies define a clear direction, foster
            collaboration between public and private sectors, and drive
            technological and systemic change.
            \item They require strong institutional support and the
            development of dynamic capabilities within government to lead
            change.
        \end{itemize}
    \end{itemize}

    \subsubsection{Analysis of Latin America's Performance}

    \noindent \textbf{Innovation (Based on Global Innovation Index 2020):}
    \begin{itemize}
        \item \textbf{Overall Performance:} The region contributes
        marginally to global innovation. The analysis covered 15 countries.
        \item \textbf{Performance by Dynamic Capability:}
        \begin{itemize}
            \item \textbf{Sensing (Weak):} The
            \textbf{institutions dimension received the lowest average score}, indicating issues with the political, regulatory, and business environment. Countries with better overall GII scores (e.g., Chile) performed better in this dimension.
            \item \textbf{Seizing (Mixed):} The
            \textbf{Business Sophistication dimension had the best average score}. However, many countries show a deficit in human capital development and RnD investment, despite having knowledge workers.
            \item \textbf{Transforming (Partial Success):} Some countries
            like Costa Rica and Mexico perform relatively well in
            transforming capabilities (creative and knowledge outputs) but
            lag in sensing and seizing, suggesting innovation occurs without
            fully considering environmental changes.
        \end{itemize}
        \item \textbf{Key Deficits:} To improve innovation, Latin America
        needs to strengthen institutions, increase investment in RnD,
        develop human capital.
    \end{itemize}

    \noindent
\textbf{Sustainable Development (Based on Sustainable Development Report 2020):}
    \begin{itemize}
        \item \textbf{Overall Performance:} Most Latin American countries
        rank in the middle portion globally. Chile, Costa Rica, Uruguay, and
        Ecuador are in the top 50 of 166 countries.
        \item \textbf{Regional Priorities (Highest-Ranked SDGs):}
        \begin{itemize}
            \item SDG 1 (No Poverty);
            \item SDG 4 (Quality Education);
            \item SDG 13 (Climate Action)
        \end{itemize}
        \item \textbf{Significant Challenges (Lowest-Ranked SDGs):}
        \begin{itemize}
            \item \textbf{SDG 9 (Industry, Innovation, and Infrastructure)}
            was the lowest-ranked goal in 10 countries.
            \item \textbf{SDG 10 (Reduced Inequality)} was the lowest-ranked
            in eight countries, indicating it is not a key priority for most
            governments in the region.
        \end{itemize}
    \end{itemize}

    \subsubsection{Proposed Roadmap and Policy Implications}\

    \noindent \textbf{Integrated Approach:} Effective strategy requires a
systemic approach that aligns strong
\textbf{institutions, mission-oriented policies, dynamic capabilities, and innovation}.\\

    \noindent \textbf{Role of Institutions:} Strong, inclusive institutions
are crucial for setting the direction through mission-oriented policies that
coordinate public and private actors. Weak institutions make it difficult to
address critical problems.\\

    \noindent \textbf{Role of Government (Public Sector):}
    \begin{itemize}
        \item Government should not just fix market failures but actively \textbf{co-create markets and ecosystems}.
        \item The public sector must take a leading role in translating
        societal challenges into concrete missions. This requires compotent
        public agencies and dynamic capabilities within the state.
    \end{itemize}

    \noindent \textbf{Human Capital:} Substantial investment in education
and human resource development is essential to build the skills required for
innovation.\\

    \noindent \textbf{Policy Focus:} Policies must encourage a shift from a
risk-management approach to one focused on seizing opportunities for
sustainable innovation.

    \subsection{Eduardo Gudynas (2013) Debates on Development and its Alternatives in Latin America: A Brief Heterodox Guide, in Beyond Development: Alternative Visions from Latin America, M. Lang and D. Mokrani, eds. Quito: Fundación Rosa Luxemburg, pp.15-39.}
    \subsubsection{The Conventional Concept of Development}

    \noindent \textbf{Definition and Origins:} Development is conventionally
understood as economic and social progress, often tied to growth,
modernization, and higher standards of living. Its modern usage in social
sciences solidified after World War II, establishing a division between
`developed' and `underdeveloped' nations.\\

    \noindent \textbf{Core Ideas (Mid-20th Century):}
    \begin{itemize}
        \item Development becasue almost synonymous with economic growth.
        \item It was viewed as a \textbf{linear process} where
        `underdeveloped' countries should emulate Western nations, as
        described in Rostow's stages of growth.
        \item Indicators like Gross Domestic Product (GDP) became primary
        targets.
        \item This model legitimized resources exploitation and the
        destruction of cultures and the environment in the name of progress.
    \end{itemize}

    \subsubsection{Early Critiques and `Development Alternatives'}

    \noindent These critiques question the \textit{methods} of devlopment
but often accept its core goal of material progress.\\

    \noindent \textbf{Dependence Theory (1960s-1970s):}
    \begin{itemize}
        \item Challenged the linear view, arguing that
        \textbf{underdevelopment is a consequence of development}, not a
        prior stage.
        \item Ascribed this condition to colonialism, imperialism, and
        asymmetrical power relations between the industrial `centre' and the
        resource-exporting `periphery'.
        \item While critical, this school still prioritized
        industrialization and economic growth.
    \end{itemize}

    \noindent \textbf{Ecological Limits to Growth (1970s):}
    \begin{itemize}
        \item The 1972 Club of Rome report, \textit{The Limits to Growth},
        argued that \textbf{perpetual economic growth is impossible} on a
        finite planet.
        \item This reporth was widely attacked, including by some Latin
        American intellectuals who saw it as a threat to modernization adn
        the use of the region's resources.
        \item A Latin American response,
        \textit{Catastropher or New Society? (1975)} argued problems were
        sociopolitical, not physical, and proposed a socialist model but
        still defended economic growth, albeit regulated.
    \end{itemize}

    \noindent \textbf{Human-Centered Approaches:}
    \begin{itemize}
        \item Sought to separate development from purely economic growth and
        focus on social dimensions.
        \item Key concepts include:
        \begin{itemize}
            \item \textbf{Another development:} Emphasized meeting basic
            needs, self-reliance, and endogeneity (defined within each society).
            \item \textbf{Human scale development (Max Neef):} Focused on
            people, not objects, and addressed poverty as a plural concept
            of unmet needs.
        \end{itemize}
    \end{itemize}

    \noindent \textbf{Sustainable Development:}
    \begin{itemize}
        \item Initially (1980), it meant using renewable resources without
        exceeding their renewal rates to meet human needs.
        \item The WCED's `Our Common Future' (1988) report offered a widely
        cited by polysemic definition that ultimately
        \textbf{reconciled sustainability with economic growth}, turning
        opposites into mutually dependent concepts.
        \item This was criticized for being a `contradiction in terms' as
        ``nothing physical can grow indefinitely''.
    \end{itemize}

    \subsubsection{The Ideology of Progress and the Post-Development Critique}

    \noindent This school of though critiques the foundational
\textit{ideas} behind development itself.\\

    \noindent \textbf{Development as Ideology:} The persistence of
development, despite its failures, suggests it functions as an
\textbf{ideology} -- the contemporary expression of the ideology of
\textbf{progress}. This ideology is shared across conventional liberal,
conservative, and socialist thought.\\

    \noindent \textbf{Post-Development:}
    \begin{itemize}
        \item A post-structuralist approach that deconstructs development
        as a \textbf{discourse} that shapes thought, institutions, and
        practices.
        \item It does not seek `another development' but rather examines its
        ecological foundations.
    \end{itemize}

    \noindent \textbf{Critique of Modernity:} Questioning development
requires questioning the project of \textbf{Modernity}, which is rooted in:
    \begin{itemize}
        \item A universalizing model based on European culture;
        \item A linear concept of history as progress from `backwardness';
        \item A sharp \textbf{duality separating society from Nature}, with
        Nature being an object for human use.
    \end{itemize}

    \noindent \textbf{Coloniality of Power:} Modernity was introduced to
Latin America via colonialism, imposing certain ideas about society,
knowledge, and history while excluding others (e.g., indigenous knowledge
systems).

    \subsubsection{Alternatives to Development}

    \noindent These frameworks seek to move beyond the ideology of progress
and the assumptions of Modernity.\\

    \noindent \textbf{Core Concepts:}
    \begin{itemize}
        \item They are \textbf{post-capitalist and post-socialist}, breaking
        with the shared ideology of progress.
        \item They challenge the society/Nature duality, often incorporating
        biocentric ethics (recognizing Nature's intrinsic view) and
        relational ontologies.
        \item They prioritize \textbf{well-being in a broad sense},
        including collective, spiritual, and ecological dimensions, over
        material accumulation.
        \item They require intercultural exchange and the inclusion of other
        knowledge systems, especially indigenous ones.
    \end{itemize}

    \noindent \textbf{Buen Vivir (`Good Living'):}
    \begin{itemize}
        \item An example of alterantive to development, drawing heavily on
        Andean indigenous knowledge (e.g., \textit{sumal kawasy, suma qamana}) as well as critical Western traditions like biocentrism and feminism.
        \item It breaks with anthropocentrism and linear progress to focus
        on community well-being in harmony with Nature.
        \item It can be understood as a political platform for building
        social orders free from the constraints of modernity.
    \end{itemize}

    \subsubsection{Contemporary Context: Progressive Governments and Neoextractivism}

    \noindent Since the late 1990s, many Latin American `progressive'
governments have rejected neoliberalism. However, in practice, they largely
adhere to a conventional development model based on
\textbf{economic growth, investment and exports}.\\

    \noindent This has led to \textbf{progressive neoextractivism}: the
intensive exploitation of natural resources (mining, oil), to fund social
programs. This model perpetuates environmental and social harm, denies
ecological limits, and reproduces core tenants of development myth
identified by Furtado decades ago.

    \section{Governance and Protest in the Twenty-First Century}

    \subsection{Rosalía Cortés (2009) Social Policy in Latin America in the Post-neoliberal Era. In: Jean Grugel and Pía Riggirozzi (eds) Governance after Neoliberalism in Latin America. Palgrave Macmillan, New York.}
    \subsubsection{Introduction to Social Policy Evolution in Latin America}

    \noindent \textbf{Focus:} Social security and labour law refors since
mid-1980s.\\

    \noindent \textbf{Guiding Argument:} Governing coalitions determine
national economic strategy, influencing service distribution and social
protection access.\\

    \noindent \textbf{Mechanisms:} State-directed redistribution via
institutions covering goods/services, social assistance, insurance, and
labour legislations. These shape social protection levels and economic
rights distributions.\\

    \noindent \textbf{Aim:} Analyze new social policies in an emergent
post-neoliberal era.\\

    \noindent \textbf{Context:} Regional social policy matrix reshaped by
economic liberalization and neoliberal growth paradigm in late 1980s-1990s.\\

    \noindent \textbf{Neoliberal Social Policy:} Result of new governing
coalitions incorporating business interests (local/global) and International
Financial Institutions (IFIs) demands.

    \subsubsection{Changing Paradigms of Social Policy}

    \noindent
\textbf{Import Substitution Industrialization (ISI) Eta (1940s-mid-1970s)}
    \begin{itemize}
        \item \textbf{Countries:} Larger Latin American countries (
        Argentina, Brazil, Chile, Mexico).
        \item \textbf{Alliances:} Strengthened political alliances between
        governing parties, employers, and trade unions.
        \item \textbf{Union Influence:} Trade unions influenced policy,
        securing social insurance benefits and income policies.
        \item \textbf{Social Policies:} Broad access in education and
        health with stable budgets and centralized implementation.
        \item \textbf{Problems:} Recurrent financing issues, cyclical
        downturns, inflation resolved by IFI-sponsored stabilization (
        wage/expenditure constraints). Led to stagnating investment (
        infrastructure, health, education) and deteriorating living
        standards (`lost decade' of the 1980s).
        \item \textbf{Neglect:} Social Policies and labour law reinforced
        government-union links, directed protection mainly to workforce,
        neglecting rural and poorest populations. This segmented provision
        created long-term difficulties for universal access to welfare and
        justice.
    \end{itemize}

    \noindent \textbf{End of ISI and Neoliberal Shift (1980s Onwards):}
    \begin{itemize}
        \item \textbf{Transition:} ISI ended in the 1980s, economic strategies
        shifted from import substitution to open economies.
        \item \textbf{Governing Coalitions:} New governments eliminated
        unions from policy debates, adopted IFI strategies, and accommodated
        local/global business groups.
        \item \textbf{Neoliberal Reforms:}
        \begin{itemize}
            \item Curtailment of public expenditure;
            \item Privatization fo social security;
            \item Introduction of labour regulation flexibilization
            \item \textbf{Official View:} Policies would protect the poor
            from economic transition impacts.
            \item \textbf{Reality:} Reinforced negative impacts of open
            markets on labour and living conditions.
        \end{itemize}
        \item \textbf{Business Dominance:} Business interest, supported by
        IFIs, pushed for reduced state economic regulation and social
        protection.
        \item \textbf{Corporatism:} In Argentina and Mexico, historic links
        between governing parties and labour helped contain union
        opposition, leading to corporatist agreements for reform and social
        security privatization.
        \item \textbf{Initial Outcomes:} Neoliberal policies and financial
        liberalization brought immediate price stabilization and short-lived
        growth (except Chile).
        \item \textbf{Negative Outcomes:} Open markets led to declining
        output, business closures, and growing unemployment. International
        financial crises (mid-1990s) exacerbated economic crisis.
        \item \textbf{Stagnation and Inequality:} Prolonged period of
        stagnation and increasing inequality (1998-2003) contributed to
        social unrest.
        \item \textbf{Social Movements:} Weakened trade unions meant social
        movements and civil society organizations represented marginalized
        groups, demanding better living standards and state intervention.
        \item \textbf{Political Shift:} Led to dramatic political changes;
        Left-leaning presidents emerged since early 2000s, reflecting
        demands for change.
        \item \textbf{Post-Neoliberal Era:} Policy mix and macroeconomic
        reshaping varies by country; attempts to combine social citizenship
        with market-led policies, or state-centered economy. No consolidated
        new paradigm, but gradual, tentative alternatives to neoliberalism.
    \end{itemize}

    \subsubsection{Economic and Social Reform Under Neoliberalism}

    \noindent \textbf{Paradigm Shift:} Economic strategies shifted from
import substitution to open economies, reflecting global trends and new
governing coalitions.\\

    \noindent \textbf{Governing Coalitions:} Eliminated trade unions from
policy debates, acquiesced to local/global business demands, adopted IFI
strategies.\\

    \noindent \textbf{Neoliberal Social Reform (Key Pillars):}
    \begin{itemize}
        \item \textbf{Curtailment of public expenditure};
        \item \textbf{Privatization of social security};
        \item \textbf{Flexibilization of labour regulations}.
    \end{itemize}

    \noindent \textbf{Official vs. Actual Impact:} Policies officially aimed
to protect the poor, but in fact reinforced negative impacts of open markets
on labour and living conditions.\\

    \noindent \textbf{Role of Business/IFIs:} Business interests (supported
by IFIs) gained dominance, pushing for reduced state regulation and social
protection.\\

    \noindent \textbf{Political Maneuvering:} In Argentina and Mexico,
corporatist agreements (historic links between parties and labour) helped
contain union opposition to reforms.\\

    \begin{itemize}
        \item [A.] \textbf{Social Security Reforms under Neoliberalism:}
        \begin{itemize}
            \item \textbf{Privatization:} Began around 1992 (Argentina,
            Bolivia, Columbia);
            \item \textbf{Shift in Responsibility:} Pensions becamse an
            individual responsibility, replacing shared risks with privately
            managed individual savings accounts.
            \item \textbf{Scheme Structure:} Typically maintained a public
            basic pensions system and introduced a mandatory, fully financed
            private scheme with fixed worked contributions and variable
            benefits.
            \item \textbf{Consequences:} Maintained or deepened earning
            inequalities and breached the social contract between generations.
            \item \textbf{Limited Initial Opposition:} Existing public
            schemes had limited coverage (excluding rural and informal
            workers), so privatization mainly affected formal sector
            workers. Public schemes wsere also seen as financially failing,
            fostering a `no-alternative' view aong the urban middle class.
            \item \textbf{IFI Influence:} IFIs favoured pension
            privatization; debate exists on whether they conditioned loans
            or indirectly influenced market-oriented models. Domestic
            support was crucial.
            \item \textbf{Country Examples:}
            \begin{itemize}
                \item \textbf{Argentina:} Introduced a \textbf{mixed system} (Reformed public basic pensions + fully funded second tier managed by unions/private banks) after intense negotiations, gaining union approval through economic benefits.
                \item \textbf{Mexico:} Initial social security (IMSS) for
                ISI-era workers faced bankruptcy by the 1980s, exacerbating
                poverty. Ministry of Finance pushed for privatization. A
                1992 scheme avoided structural change, but the 1995
                financial crisis led to a new 1997 scheme: increased
                retirement benefit threshold, closed the old scheme for new
                entrants (who joined private schemes), while maintaining
                privileged pensions for state officials. State's financing
                shared increased, eventually shifting all formal labour to
                the new system.
                \item \textbf{Brazil:} Did not follow the region's
                privatization pattern due to its segmented labour market.
                The 1988 Constitutions recognized universal social rights,
                but 1990s economic problems limited expansion. President
                Cardoso made small changes for private sector workers.
                Brazil implemented a `vertical mass expansion', extending
                targeted social assistance to uninsured workers, rather than
                universal access.
            \end{itemize}
            \item \textbf{Outcomes:} Low coverage persisted (2005: 65\%
            formal, 21\% informal workers contributed). Compliance fell.
            Private management became concentrated, with high administrative
            costs.
        \end{itemize}
        \item[B.] \textbf{Labour Market Reforms under Neoliberalism:}
        \begin{itemize}
            \item \textbf{Traditional Context:} Labour law was generally
            protective, guaranteeing employment regularity for formal
            workers (e.g., penalizing dismissals, limiting fixed-term
            contracts).
            \item \textbf{Neoliberal Reforms:} Aimed to flexibilize
            contracts, reduce termination penalties, facilitate outsourcing,
            and link wage increases to productivity.
            \item \textbf{Impact:} Negatively affected workers' living
            standards and increased insecurity.
            \item \textbf{Country Examples:}
            \begin{itemize}
                \item \textbf{Argentina:} Deep reforms, relaxing dismissal
                costs and transferring employment responsibilities to
                state/agencies/subcontractors, with some union support.
                \item \textbf{Mexico:} Adapted labour costs/discipline
                without new legislation. Agreements with trade unions (CTM)
                allowed flexible working, short-term contracts, and
                subcontracting. Fixed earnings meant real wages dropped 60\%
                over 20 years due to inflation.
                \item \textbf{Brazil:} 1988 Constitution increased worker
                protections. President Cardoso's 1998 legislation increased
                flexible working, limited individual rights, introduced
                part-time contracts, and removed fixed-iterm contract limits.
            \end{itemize}
            \item \textbf{Overall Outcomes:} Curtailed individual worker
            rights, expanded short-term contracts, leading to increased
            employment and income security. The number of unregistered
            workers (without basic labour rights) grew.
            \item \textbf{IFI Response:} IADB and World Bank programs
            provided loans for technical cooperation, training, and public
            works to manage reform fallout and embed labour market reforms.
        \end{itemize}
    \end{itemize}

    \subsubsection{Legacy of Economic Social Reform (Neoliberalism)}

    \noindent \textbf{Economic Performance:} Improved export performance,
controlled inflation, reduced fiscal deficits. However, output growth fell (
from 5.7\% to 3.5\% annually). Financial liberalization increased
vulnerability to foreign funds.\\

    \noindent \textbf{Shrinking State Protection:} State intervention
reoriented towards targeted assistance, reducing social protection and
enhancing the labour market's role in shaping living conditions.\\

    \noindent \textbf{Debate on Social Inequality:}
    \begin{itemize}
        \item \textbf{Walton (2004):} Neoliberal transformation `highly
        beneficial', increased growth without increasing social inequality.
        Attributed social problems to preexisting inequalities/ineffective
        institutions, nort market reforms. Noted mixed imapcts on
        inequality (Argentina up, Mexico down, Brazil improved).
        \item \textbf{Huber and Solt (2004):} Market reforms yielded
        disappointing results for social equity and democracy.
        \item \textbf{Contradictory Data:} Labour market and poverty data
        often contradict Walton's view.
        \begin{itemize}
            \item \textbf{Employment:} Unemployment rose (8.2\% to 10.5\%
            1990-2000), informal employment grew due to manufacturing
            crisis, privatization, and public sector cuts. Public sector
            jobs declined.
            \item \textbf{Job Quality/Security:} Deteriorated, leading to
            employment instability and income security.
            \item \textbf{Inequality:} Increased income concentration and
            widening inequality, with wage differentials between
            formal/informal sectors and declining informal sector incomes.
            Income share of poorest 40\% fell to 10\%.
            \item \textbf{Poverty:} While poverty dropped early 1990s, by
            1999 it was still higher than in the 1980s, and the income of
            the poor increased less than the non-poor.
        \end{itemize}
    \end{itemize}

    \noindent \textbf{Persistent Inequality:} High poverty levels attibuted
to persistent long-term income and asset inequality.\\

    \noindent \textbf{Social Services:} Changes reinforced barriers to
accessing social services. Administrative reforms insufficient; added
financial burden for the poor Middle/high-income groups shifted to private
providers.
    \begin{itemize}
        \item \textbf{Brazil (Exception):} Broadened social/political access
        to services to some extent, but mainly expanded social security
        within the formal sector, leaving informal workers ``out of the
        system''.
    \end{itemize}

    \noindent \textbf{Civil Society and Clientelism:} Claims that civil
society channeled demands of the poor, unlike political parties. However,
targeted local programs often reproduced exclusionary practices and
facilitated clientelism. Local political leaders controlled beneficiary
selection, reinforcing dependency. Not clear if it led to greater voice for
the poor.\\

    \noindent \textbf{Overall Conclusion of Neoliberal Era:} Economic
reforms transformed the social world, with social policy reinforcing (not
mediating) economic changes. Weakened labour movements and growing civil
society occurred amidst recession, falling incomes, and shrinking social
protection, leading ot policy crossroads in the early 21st century.

    \subsubsection{Emergence of New Economic and Social Strategies (Post-Neoliberal Era)}

    \noindent \textbf{Context (Post-2003):} Favourable global economy (
commodity prices, exports) improved fiscal accounts. Crisis of 1998-2003
eroded previous coalitions, leading to Left-leaning government winning
elections (Argentina, Brazil, Venezuela, Bolivia, Chile) on anti-neoliberal
platforms.\\

    \noindent \textbf{Uncertainty:} Unclear how much these new governments
will dismantle existing neoliberal programs or create entirely new strategies.\\

    \noindent \textbf{Country-Specific Trends:}
    \begin{itemize}
        \item \textbf{Argentina:}
        \begin{itemize}
            \item Began labour market re-regulation (2004).
            \item Reformed retirement system, encouraging return to public
            system, extended non-contributory pensions.
            \item Reoriented economic strategy (2002): currency devaluation
            boosted manufacturing/domestic market, urban employment.
            \item \textbf{Reinstated Collective Bargaining,} revived
            tripartite council for minimum wages (tripled by 2008).
            \item \textbf{Limitations:} Income policy priarily benefits
            formal waged workers, neglecting informal/precarious jobs.
            \item \textbf{Targeted Programs:} ``Plan Jefes y Jefas de Hogar
            Desocupados'' (cash transfers for household heads with conditons
            like schooling/vaccination) had broader but still insufficient
            coverage, with benefit values eroded by infaltions.
            \item \textbf{Emerging Model:} Based on a
            \textbf{dichotomy enhancing rights for formal sector workers while neglecting others}, intensifying the formal/informal labour divide. Shift from `poverty' to `labour issues' on policy agenda.
        \end{itemize}
        \item \textbf{Brazil:}
        \begin{itemize}
            \item \textbf{Lula da Silva:} Continued social security
            expansion for formal sector employees while directing expanding
            to poor families (``moderate reformism''). Unified
            public/private retirement schemes, reduced public employee benefits.
            \item \textbf{Targeted Programs:} Expanded conditional cash
            transfer programs (e.g., Bolsa Familia, which absorbed Fome Zero)
            . Bolsa Familia covered 11 million families by 2006, aiming to
            reduce poverty and incentivize school/health access.
            \item \textbf{Critique:} Coverage expansion may have political
            motivations.
            \item \textbf{Overall:} Combines neoliberal continuity in social
            security with new attention to targeted programs, increased
            minimum wages, and federal centralization of social policy. The
            overall policy thrust remains largely inherited.
        \end{itemize}
        \item \textbf{Mexico:}
        \begin{itemize}
            \item Social policy continued towards assistance for the
            vulnerable/poor, away from labour market support. Reflects
            collapse of ISI social coalition and declining importance of
            social security/labour rights.
            \item Prioritizes market role, with interventions restricted to
            cash transfer programs for the extremely poor.
            \item \textbf{Pioneer in CCTs:} Progresa (later Oportunidades)
            alleviated extreme poverty through conditional cash transfers (
            health checks, school attendance). Reached 24\% of the
            population by 2005, showing some success in secondary school
            enrollment and child nutrition.
            \item \textbf{Limitations:} Less effective for primary school
            attendance; doesn't solve long-term access to health/education (
            90\% of poor lacked healthcare in 2006).
            \item \textbf{Overall:} Fragmented social policies, with
            components from both pre- and post-neoliberal paradigms. Social
            security not fully privatized, education/health decentralized.
            Focus shifted from workers to extreme poverty.
        \end{itemize}
    \end{itemize}

    \subsubsection{Current Status of Social Policy in Latin America}

    \noindent \textbf{Cash Transfer Programs (CTPs):} Remain the
\textbf{paradigm of social policy} in Latin America, largely due to
governmental commitment to fiscal equilibrium.\\

    \noindent \textbf{Limitations of CTPS:}
    \begin{itemize}
        \item \textbf{Partial Coverage:} Only provide partial coverage and
        satisfy basic needs, especially without broader redistributive
        measures (e.g., tax reform). Exclude important segments of the
        poor (e.g., poorly paid workers).
        \item \textbf{Bypass Institutions:} Often bypass public social
        institutions, rarely building independent citizenship.
        \item \textbf{Social Control/Clientelism:} Can facilitate
        interaction between governments and the poor, potentially serving as
        a means fo social control or populist strategies by co-opting the
        poort and CSOs responsible for distribution. Clientelism arises when
        actors have discretion over benefit allocation.
    \end{itemize}

    \noindent \textbf{Overall Emerging Model:} A
\textbf{segmented and limited model}, divided between regulations/provisions
for formal workers and cash transfer programs for ``the poor''.\\

    \noindent \textbf{Future Challenges:} Universal social security
coverage, full access to quality employment, and improved education/health
are generally \textbf{not yet on the agenda}.\\

    \noindent \textbf{Regional Differences:} Argentina has begun to undo
elements of the neoliberal legacy, but this is less evident in Mexico,
Brazil and Chile

    \subsection{Marie-Christine Doran (2017) The Hidden Face of Violence in Latin America: Assessing the Criminalization of Protest in Comparative Perspective. Latin American Perspectives 44(216): 183-2016.}

    \subsubsection{Defining Criminalization of Protest}
    \noindent \textbf{Terminology:} Also referred to as criminalization of:
    \begin{itemize}
        \item Collective action/social protest;
        \item dissent;
        \item citizen participation;
        \item human rights defense.
    \end{itemize}

    \noindent \textbf{Nature:}
    \begin{itemize}
        \item Goes beyond mere penalization (imprisonment, new laws against
        activism).
        \item Invovles \textbf{intimidation and repression} based on
        antagonizing civil/political rights (essential to democracy) as
        threats to national security/interest.
        \item \textbf{Delegitimizes citizens} and paves the way for
        unpunished human rights violations.
        \item Analytically, it's a \textbf{collective trend} that takes into
        account power relations and critically analyzes the discursive
        battle over what is `peaceful' vs. `illegal' or `violent'.
        \item Includes enacting restrictive legislation, impunity for state
        officials, and refusal to investigate violations.
        \item Transforms peaceful, democratic actions into crimes subject to
        punishment.
        \item Legitimizes repression, turning community members into `public
        enemies' accused of violence, delinquency, or terrorism.
    \end{itemize}

    \textbf{Purpose:}
    \begin{itemize}
        \item \textbf{Obscures the political nature} of violence, presenting
        it as an objective fact necessary for security.
        \item A form of \textbf{depoliticization}, neutralizing political
        aspects and providing legal cover fore economic decisions.
        \item A indirect attack on the legitimacy of basic democratic conduct,
        leading to disregard for human life and gross human rights violations.
        \item Aims to \textbf{restrict democracy} and make it compatible
        with routinized violence against opponents, leading to
        ``violence-compatible democracies'' or ``violent pluralism''.
    \end{itemize}

    \subsubsection{Causes and Enabling Factors of Criminalization}

    \noindent Criminalization is rooted in:
    \begin{itemize}
        \item [$1$.] \textbf{Continuity with Authoritarianism:}
        \begin{itemize}
            \item \textbf{Identification of an internal threat} justifying
            unprecedented security policies.
            \item Militarization of police and increased military control
            over politics.
            \item Security policies in Latin America are continuous with
            antisubversion policies of authoritarian regimes.
        \end{itemize}
        \item[$2$.] \textbf{Influence of Low-Intensity Democracy:}
        \begin{itemize}
            \item \textbf{Reduction of democracy to a purely formal concept} (e.g., free elections, multi-party systems).
            \item \textbf{Delegitimization of all claim-based conflict}.
            \item This narrow view allows measures that render previously
            peaceful and legitimate practices illegal.
            \item Targets protests against natural-resource-based industry,
            educational rights, union rights, indigenous rights, and
            struggles against impunity.
        \end{itemize}
        \item[$3$.]
        \textbf{``Blaming the Victims'' under National Reconciliation:}
        \begin{itemize}
            \item Strategy to eliminate conflict and avoid political
            instability, or a new `breakdown of democracy'.
            \item State violence is attributed to the people's actions,
            making victims of dictatorship or state violence illegitimate.
            \item Justice for human rights violations is precluded, often
            justified as ``disadvantageous to democratic processes''.
            Impunity enshrined in reconciliation processes weakens the rule
            of law.
            \item Fear of social conflict and demands (especially for
            justice) has limited participation and led to regimes
            characterized as `post-dictatorships'.
        \end{itemize}
        \item[$4$.] \textbf{Intensification of Security Policies:}
        \begin{itemize}
            \item Latin America, identified as the most violent continent,
            has become a ``laboratory for security policies'' (war on
            terrorism, drug trafficking, crime).
            \item These policies create new fronts against social conflict,
            where any opposition is deemed a threat.
            \item The state and police adopt a new role: protecting
            citizens (seen as victims of themselves) from uncontrollable
            violence, justifying a strong, Hobbesian state.
            \item This disclosure reduces past state responsibility for
            violence and reasserts state control over a ``violence-prine
            society''. Militarization of society, as seen in Mexico, expands
            the military's role into civil society segments.
        \end{itemize}
    \end{itemize}



\end{document}